# (WIP) Emotion

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  out.width = "100%"
)
library(tidyverse)
```

## (WIP) Introduction to emotion and music

## (WIP) Modelling emotion

```{r, echo = FALSE}
DiagrammeR::grViz("
digraph boxes_and_circles {

  graph [overlap = true, fontsize = 10, layout = dot, rankdir = LR]

  
  # L1 nodes
  node [shape = box, fixedsize = true,
        width = 2.2]
  
  Happy
  Amazed
  Dazzled
  Allured 
  Moved 
  
  Inspired
  FeelTranscendence [ label = 'Feeling of transcendence' ]
  Spirituality [label = 'Feeling of spirituality']
  Thrills
  
  Love [ label = 'In love' ]
  Affectionate
  Sensual
  Tender
  Soft [ label = 'Softened-up' ]
  
  Sentimental
  Dreamy
  Nostalgic
  Melancholic
  
  Calm
  Relaxed
  Serene
  Soothed
  Meditative
  
  Energetic
  Triumphant
  Fiery
  Strong
  Heroic
  
  Stimulated
  Joyful
  Animated
  Dancing
  Amused
  
  Agitated
  Nervous
  Tense
  Impatient
  Irritated
  
  Sad
  Sorrowful

  # L2 nodes

  node [shape = ellipse,
        fixedsize = true,
        width = 2.5, 
        height = 1,
        fontsize=23] 
  Wonder
  Transcendence
  Tenderness
  Nostalgia
  Peacefulness
  Power
  JoyfulActivation [label = 'Joyful activation' ]
  Tension
  Sadness
  
  # L3 nodes
  
    node [shape = circle,
        fixedsize = true,
        width = 2.5, 
        fontsize=30] 
  
  Sublimity
  Vitality
  Unease

  
  edge [ dir = back ]
  
  # L1 edges
  
  Happy->Wonder [ label = 1.00 ]
  Amazed->Wonder [label = 0.95 ]
  Dazzled->Wonder [label = 0.84 ]
  Allured->Wonder [label = 0.86 ]
  Moved->Wonder [label = 0.75 ]
  
  Inspired->Transcendence [ label = 1.00 ]
  FeelTranscendence->Transcendence [ label = 0.92 ]
  Spirituality->Transcendence [ label = 0.90 ]
  Thrills->Transcendence [ label = 0.65 ]
  
  Love->Tenderness [ label = 1.00 ]
  Affectionate->Tenderness [ label = 0.97 ]
  Sensual->Tenderness [ label = 0.98 ]
  Tender->Tenderness [ label = 0.97 ]
  Soft->Tenderness [ label = 0.74 ]
  
  Sentimental->Nostalgia [ label = 1.00 ]
  Dreamy->Nostalgia [ label = 0.77 ]
  Nostalgic->Nostalgia [label = 0.64 ]
  Melancholic->Nostalgia [label = 0.52 ]
  
  Calm->Peacefulness [ label = 1.00]
  Relaxed->Peacefulness [ label = 0.96 ]
  Serene->Peacefulness [ label = 0.94 ]
  Soothed->Peacefulness [ label = 0.90 ]
  Meditative->Peacefulness [ label = 0.58 ]
  
  Energetic->Power [ label = 1.00]
  Triumphant->Power [ label = 0.76 ]
  Fiery->Power [ label = 0.72 ]
  Strong->Power [ label = 0.70 ]
  Heroic->Power [ label = 0.56 ]
  
  Stimulated->JoyfulActivation [ label = 1.00 ]
  Joyful->JoyfulActivation [ label = 0.99 ]
  Animated->JoyfulActivation [ label = 0.95 ]
  Dancing->JoyfulActivation [ label = 0.72 ]
  Amused->JoyfulActivation [ label = 0.56 ]
  
  Agitated->Tension [ label = 1.00 ]
  Nervous->Tension [ label = 0.85 ]
  Tense->Tension [ label = 0.63 ]
  Impatient->Tension [ label = 0.49 ]
  Irritated->Tension [ label = 0.39 ]
  
  Sad->Sadness [ label = 1.00 ]
  Sorrowful->Sadness [ label = 0.82 ]
  
  # L2 edges
  Wonder->Sublimity [ label = 1.00 ]
  Transcendence->Sublimity [ label = 0.66 ]
  Tenderness->Sublimity [ label = 0.75 ]
  Nostalgia->Sublimity [ label = 0.89 ]
  Peacefulness->Sublimity [ label = 0.65 ]
  
  Power->Vitality [ label = 1.00 ]
  JoyfulActivation->Vitality [ label = 0.95 ]
  
  Tension->Unease [ label = 1.00 ]
  Sadness->Unease [ label = 0.27 ]
  
}
") %>% DiagrammeRsvg::export_svg () %>% charToRaw %>% rsvg::rsvg_svg("images/GEMS.svg")

knitr::include_graphics("images/GEMS.svg")
```

## Measuring emotional responses

This section introduces various ways that we can measure emotional responses in the context of psychological experiments. Broadly speaking, we can organise these measurement techniques into three primary categories: self-report measures, physiological measures, and neural measures.

### Self-report

The idea behind self-report measures is simply to ask the participant about their emotional experiences. This relies on the fact that, at least in Western society, most adult participants have fairly sophisticated vocabularies for describing their emotional experiences.

There are many ways in which we could ask someone about their emotional experiences, in more or less structured ways. A particularly common approach though is to use rating scales; here we simply ask people to describe an emotional experience by rating it according to various criteria, for example 'joyfulness' or 'anger'. These numbers can then be aggregated to provide a quantitative summary of the participant's emotional responses.

There are various kinds of rating scales we could use here. One prominent option is the Geneva Emotional Music Scale, or 'GEMS' (Zentner et al., 2008), which we discuss in Section \@ref(wip-modelling-emotion). 

The GEMS takes many different emotion terms (on the left hand side here), and clusters them together into a three-level hierarchy. The scale was constructed in a data-driven way: the emotion terms are terms that frequently occur when people describe music-induced emotional experiences, and the higher-level clusters (or 'factors') describe how these different emotion terms tend to co-occur with each other.

We can use the GEMS to make questionnaires for quantifying certain kinds of core emotional responses to music. For example, suppose I want to quantify the tension that someone experiences when they listen to a particular piece of music. The GEMS tells me that a good way to do that is to give the participant rating scales for five adjectives: agitated, nervous, tense, impatient, and irritated. I get the participant to rate their experience on these five adjectives. I then combine their results, for example by averaging, and this gives me an overall 'tension' score. Similarly, the GEMS tells me that I could compute scores for a general 'unease' factor by first computing scores for 'tension' and 'sadness', and then combining them.

\

## (WIP) Mechanisms for inducing musical emotion

<!-- <!-- Can we include some music therapy here? -->

-->

<!-- Credit: Internet Archive Book Images, no restrictions, via Wikimedia Commons -->

<!-- from here: <https://docs.google.com/document/d/1arLtzHGsrLomU8mlCHXJuyYk5zfq9FPzFeKh8k8MvJk/edit#> -->

<!-- see images directory -->

<!-- #### Heart rate -->

<!-- Heart rate can be measured with inexpensive and unobtrusive heart rate monitors. Many smart watches nowadays even include heart rate monitors as standard. -->

<!-- #### Respiration rate -->

<!-- The respiration rate corresponds to the number of breaths that the participant takes per minute. One way of measuring this is with a 'pneumograph', as displayed here: a pneumograph measures respiration by tracking chest movements. -->

<!-- #### Skin conductance -->

<!-- #### Skin temperature -->

<!-- #### Facial electromyography -->

<!-- #### Pupillometry -->

<!-- Goosebumps -->

<!-- Intense emotional experiences, especially in the context of music, are often accompanied by skin tingling and goosebumps, corresponding to a process called 'piloerection' where body hairs stand on end. These experiences are called 'chills'. It's possible to record the development of goosebumps using custom cameras placed for example on the participant's arm. -->

<!-- Schematic illustration of piloerection. Credit: AnthonyCaccese, CC BY-SA 4.0, via Wikimedia Commons -->
