# Quantitative (I): Design

```{r, include = FALSE}
library(tidyverse)
library(ggpubr)
theme_set(theme_pubr())
knitr::opts_chunk$set(echo = FALSE)

```

Quantitative methods, at their core, are all about characterising two things:

(a) distributions of variables;

(b) relationships between variables.

These relationships can be summarised either numerically (e.g. using tables) or graphically (e.g. using plots). For example, we might study how fuel efficiency (measured in miles per gallon) is distributed in a collection of cars:

```{r}
mtcars |> 
  ggplot(aes(mpg)) + 
  geom_histogram(bins = 20) +
  scale_x_continuous("Fuel efficiency (miles per gallon)") + 
  scale_y_continuous("Count")
```

We might follow that up by studying how fuel efficiency relates to the car engine's horsepower:

```{r}
mtcars |> 
  ggplot(aes(hp, mpg)) + 
  geom_point() + 
  scale_x_continuous("Engine horsepower") +
  scale_y_continuous("Fuel efficiency (miles per gallon)") 
```

It feels very natural to speak about car engines in these numeric constructs such as miles per gallon and horsepower, and to quantify relationships between them using data plots. In contrast, it is less obvious how to construct equivalent numeric constructs for human psychology, where many of the relevant concepts (e.g. personality, attention, consciousness) are hard to define and even harder to measure. This task of defining appropriate variables is one of the key challenges of psychology research.

## Defining variables

### Taxonomies of variables

We can define a variable as a measurement with a well-defined set of potential values. One example might be age (in years), corresponding to a number ranging typically between 0 and 100. Another example might be nationality, which might take values such as French, German, or Italian.

Variables come in many forms, and these forms have important implications for how you design an experiment as well as how you analyse the resulting data. Here are several types of variable, as commonly defined in methods textbooks:

-   **Categorical variables (a.k.a. nominal variables).** These take one of a finite set of values, with these values having no intrinsic ordering. An example might be nationality, which could take values such as French, German, and Indian. A special kind of categorical variable is a **binary variable (a.k.a. dichotomous variable)**: this is simply a categorical variable with only two possible values, which comes up for example in True/False questions.

-   **Ordinal variables.** Like categorical variables, these take one of a finite set of values, but unlike categorical variables, these values have an intrinsic order. An example of an ordinal variable might be a participant's response to the following question: "How much do you like jazz music?" where the participant has to choose from four options labelled 1-4:

    1.  Not at all;

    2.  A little;

    3.  A lot;

    4.  Very much.

-   **Continuous variables**. These are variables that take numeric values within a continuous range. A truly continuous variable should be able to be specified to an arbitrary degree of precision: for example, we might say that the temperature is 45.832 degrees Centigrade. There are two main types of continuous variable:

    -   **Interval variables**. A variable is an 'interval' variable if (and only if) addition and subtraction are 'meaningful' operations. For example, the time of day is an interval variable, because it makes sense to say 'ten minutes after' (addition) or 'five minutes before' (subtraction) a given time.

    -   **Ratio variables.** These are like interval variables, but they have an additional property: multiplication is also a meaningful operation. Money is an example of a ratio variable, because it makes sense to say that Anna (for example) has twice as much money as Bob. Temperature in Centigrade is meanwhile **not** a ratio variable, because it doesn't make sense to say that 20°C is 20 times as hot as 1°C (at least, not if we listen to our physics teacher).

It is worth making a few notes at this point:

-   In practice, most analysis methods used in psychology do not differentiate between interval and ratio variables, so it is rarely worth losing much sleep over this distinction.

-   There are nonetheless often grey areas where different variable categories overlap. For example, a student might take a multiple-choice exam with a 100 questions, and therefore possible marks would be 1, 2, 3, ..., 99, or 100. This is technically not a fully continuous variable, because the set of possible marks is finite, and it is not possible to get 'in-between' marks such as 22.5 or 33.7. However, in many situations (e.g. data analysis) it will be useful here to treat the variable as if it is indeed continuous.

-   A particularly debated area is multiple-choice questionnaires, where participants have to choose from multiple numbered options arranged in a scale, for example:

    1.  Not at all;

    2.  A little;

    3.  A lot;

    4.  Very much.

    Some might argue that this variable is ordinal, because it's not obvious that addition and subtraction make sense here. Others might say that participants naturally understand that the scale is meant to be interpreted as an interval scale, and hence respond in a manner that is consistent with that. This question is not really resolved in the literature, but it is fairly accepted practice in these scenarios to pretend like the variable is an interval variable.

What kinds of variables might we have in music psychology experiments? There are many possibilities. Most can usually be categorised as either *participant variables* or *stimulus variables*.

-   **Participant variables.** These are properties of the participant. Many of these are personal traits that existed before the experiment, for example age, gender, prior musical training, and so on. Other of these might be variables that are determined during the experiment, for example when the experimenter assigns the participants to different experimental conditions. For example, suppose that we conduct a study on schoolchildren, investigating the effect of music lessons on intelligence; we might end up with the following table, where each row is a participant, and each column is a participant variable:

+----------+----------+-----------+---------------+----------+----------+
| Name     | Gender   | Start age | Condition     | Start IQ | End IQ   |
+==========+==========+===========+===============+==========+==========+
| Oliver   | Male     | 12        | No lessons    | 110      | 111      |
+----------+----------+-----------+---------------+----------+----------+
| George   | Male     | 11        | Music lessons | 105      | 110      |
+----------+----------+-----------+---------------+----------+----------+
| Olivia   | Female   | 11        | No lessons    | 99       | 98       |
+----------+----------+-----------+---------------+----------+----------+
| Amelia   | Female   | 12        | Music lessons | 103      | 104      |
+----------+----------+-----------+---------------+----------+----------+

: Fictional dataset illustrating different participant variables.

-   **Stimulus variables.** These are properties of a stimulus. A given music psychology experiment might involve many different musical stimuli, and we might be interested in how different stimuli elicit different responses from participants. Below is an excerpt from a real published dataset [@Bowling2018-hz] of 'attractiveness' ratings for musical chords, supplemented with computational analyses of these chords from a later publication [@Harrison2020-gx]:

```{r}

readRDS("input/consonance/perception-bowling.rds") %>% 
  transmute(
    # Chord = Hmisc::capitalize(name), 
    `Interval (semitones)` = map_chr(pi_chord, ~ paste(sprintf("%.2f", diff(.)), collapse = ", ")),
    # Pitches = map_chr(pi_chord, ~ paste(sprintf("%.2f", .), collapse = ", ")),
    `Attractiveness rating` = sprintf("%.2f", rating),
    Roughness = sprintf("%.2f", hutch_78_roughness),
    Harmonicity = sprintf("%.2f", har_18_harmonicity),
    Familiarity = sprintf("%.2f", - scale(har_19_corpus))
  ) %>% 
  slice(1:12) %>% 
  knitr::kable(caption = "Attractiveness ratings for dyads from @Bowling2018-hz supplemented with computational analyses from @Harrison2020-gx.")
```

### The challenge of psychological measurement

It is relatively easy to define a valid measurement process for many physical variables, for example height and weight. You can measure someone's height with a tape measure, and you can measure their weight with a weighing scale. Sure, if you're doing it properly you might have to think carefully about some subtle points of the methodology (should the height measurement include the person's hair? does it matter whether we measure the person in the morning versus the evening, or before or after a large meal?), but conceptually it is not too difficult to convince oneself of the validity of the measurement process.

In contrast, it is relatively complicated to define a valid measurement process for many psychological variables, because these variables are generally much more theoretical in their conceptualisation. Take *personality* as an example. Psychologists often conceptualise personality using the so-called *Big Five* model, which characterises personality in terms of five underlying factors:

1.  Extraversion;

2.  Agreeableness;

3.  Openness to experience;

4.  Conscientiousness;

5.  Neuroticism.

The idea is that each person's personality can be summarised (to some extent) as a collection of five scores, one for each of these five dimensions.

you might have some questions

Physical variables such as height and weight

## Studying relationships between variables

-   Surveys

-   Behaviour (what do people do?)

    -   Reaction times

    -   Choices (gambling, music listening)

-   Neuroimaging

Relationships between variables What is a variable? (measurement, operationalisation) (Kinds of variables) What can we read into a relationship? (association, prediction, causality) How reliable is a relationship? (inference) How generalisable is a relationship? (generalisability)

## Measurement

## Causality

## Generalisability

## Ethics

## Examples

#### Psychological studies

#### Corpus studies

#### Acoustic studies
